Learning to Pivot with Adversarial Networks

### 2017-05-18 tex/nips/nips_2017.tex
bring in Shimmin into experiments
Had to chop some sentences to make it fit.

\author{
  Gilles Louppe\\
  New York Univeristy\\
  \texttt{g.louppe@nyu.edu}\\
  \And
  Michael Kagan\\
  SLAC National Accelerator Laboratory\\
  \texttt{makagan@slac.stanford.edu}\\
  \And
  Kyle Cranmer\\
  New York Univeristy\\
  \texttt{kyle.cranmer@nyu.edu}
}

@@

\begin{abstract}
    % Many inference problems involve data generation processes that are not uniquely
    % specified or are uncertain in some way.
    In a scientific context, the presence of
    several plausible data generation processes is often associated to systematic uncertainties. Robust inference is possible if it is based on a
    pivot -- a quantity whose distribution does not depend on the unknown value of the
    (typically continuous) nuisance parameters that parametrizes this family of
    generation processes. In this work,  we introduce a training procedure
    based on adversarial networks for enforcing the pivotal property on a predictive model.
    We derive theoretical results showing that the procedure tends
    towards a minimax solution corresponding to a predictive model that is both
    optimal and independent of the nuisance parameters (if that model exists) or
    for which one can tune a trade-off between accuracy and robustness. We
    demonstrate the effectiveness of this approach with a toy example and an example
    from particle physics.
\end{abstract}

@@

    \item $\smash{\hat\theta}_f$ maximizes the conditional entropy
    $H({Z|f(X;\theta_f)})$, since $H(Z|f(X;\theta)) \leq H(Z)$ from the properties of entropy. Note that this
    latter inequality holds for both the discrete and the differential definitions of entropy.
\end{itemize}
By assumption, the lower bound is active, thus we have $H(Z|f(X;\theta_f)) = H(Z)$
because of the second condition, which happens exactly when $Z$ and $f(X;\theta_f)$
are independent variables. In other words,  the
optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is also a pivotal
quantity.

\end{proof}

@@

\subsection{Toy example}
\label{sec:toy}

@@

\begin{figure}
\centering
\begin{minipage}{.46\linewidth}
    \centering
        \includegraphics[width=\textwidth]{../figures/training.pdf}
        \caption{Toy example. Training curves for ${\cal L}_f(\theta_f)$, ${\cal L}_r(\theta_f, \theta_r)$
                 and ${\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r)$.
                 Initialized with a pre-trained classifier $f$, adversarial training was performed for $200$ iterations, mini-batches of size $M=128$, $K=500$ and $\lambda=50$.}
        \label{fig:toy-training}
\end{minipage}
\hspace{.05\linewidth}
\begin{minipage}{.46\linewidth}
    \centering
        \includegraphics[width=\textwidth]{../figures/ams.pdf}
        \caption{Physics example. Approximate median significance as a function of the decision threshold
                 on the output of $f$. At $\lambda=10$, trading
                 accuracy for independence to pileup
                 results in a net benefit in terms of statistical significance.}
        \label{fig:physics-ams}
\end{minipage}
\end{figure}

\subsection{High energy physics example}
\label{sec:hep}

Experiments at high energy colliders like the LHC \citep{LHCMachine} are
searching for evidence of new particles beyond those described by the
Standard Model (SM) of particle physics. A wide array of
theories predict the existence of new massive particles that would decay to
known particles in the SM such as the $W$ boson. The $W$ boson is unstable and
can decay to two quarks, each of which produce collimated sprays of particles
known as jets.
%(collectively denoted $W\to jj$).
If the exotic particle
is heavy, then the $W$ boson will be moving very fast, and  relativistic effects
will cause the two jets from its decay to merge into a single `$W$-jet'. These
$W$-jets have a rich internal substructure (see e.g.
\citep{Altheimer:2012mn,Altheimer:2013yza}). However,
jets are also produced ubiquitously at high energy colliders through more
mundane processes in the SM, which leads to a challenging classification problem
that is beset with a number of sources of systematic uncertainty.
The classification challenge used here is common in jet substructure
studies (see e.g. \citep{Khachatryan:2014vla,ATL-PHYS-PUB-2015-033,wbosonATLAS}): we aim to distinguish normal jets produced copiously
at the LHC ($Y=0$) and from $W$-jets ($Y=1$) potentially coming from an exotic process.
We reuse the datasets used in
\citep{baldi2016jet}.
% For completeness, the jets are built with the
% anti-$k_T$ algorithm \citep{Cacciari:2008gp}
% with radius parameter $R=1.2$, and trimmed \citep{Krohn:2009th} with subjets built with
% the $k_T$ algorithm and parameter $f_{cut}=0.2$. The features used
% in the classifiers are: trimmed jet invariant mass, N-subjettiness
% $\tau_{21}^{\beta=1}$ \citep{nsub,Thaler2012}, and the energy correlation
% functions  $C_{2}^{\beta=1}$, $C_{2}^{\beta=2}$,
% $D_{2}^{\beta=1}$, $D_{2}^{\beta=2}$ \citep{Larkoski2013}.

@@

We do not expect that we will be able to find a function $f$ that simultaneously
minimizes the classification loss ${\cal L}_f$ and is pivotal. Thus, we need to optimize
the hyper-parameter $\lambda$ of Eqn.~\ref{eqn:vf-lambda} with respect to
a higher-level objective. In this case, the natural higher-level context is a
hypothesis test of a null hypothesis with no $Y=1$ events against an
alternate hypothesis that is a mixture of $Y=0$ and $Y=1$ events.
In the absence of systematic uncertainties, optimizing ${\cal L}_f$ simultaneously
optimizes the power of a classical hypothesis test in the Neyman-Pearson sense.
When we include systematic uncertainties we need to balance the
classification performance against the robustness to uncertainty in $Z$.
Since we are still performing a hypothesis test against the null, we only
wish to impose the pivotal property on $Y=0$ events. To this end,
we use as a higher level objective the Approximate Median Significance (AMS), which is a natural generalization
of the power of a hypothesis test when systematic uncertainties are taken into account
(see Eqn.~20 of \citep{adam2014higgs}).

@@

@@

While our contribution may first appear as a minor extension, most previous
works are based on the strong and limiting assumption that $Z$ is a binary
random variable (e.g., $Z=0$ for the source domain, and
$Z=1$ for the target domain). In particular, \citep{pan2011domain,gong2013connecting,baktashmotlagh2013unsupervised,zemel2013learning,ganin2014unsupervised,ajakan2014domain,EdwardsS15,louizos2015variational}
are all based on the minimization of some form of divergence between the two distributions of
$f(X)|Z=0$ and $f(X)|Z=1$. For this reason, and in contrast with our procedure,
these works cannot intrinsically be generalized to non-binary or
continuous nuisance parameters, both from a practical and theoretical point of view.
Of notable interest however, \cite{kamishima2012fairness} enforces fairness through
a prejudice regularization term based on empirical estimates of $p(f(X)|Z)$.
While this approach is in principle sufficient for handling non-binary
nuisance parameters $Z$, it requires accurate
empirical estimates of $p(f(X)|Z=z)$ for all values $z$, which quickly becomes
impractical as the cardinality of $Z$ increases. By contrast, our approach models the conditional
dependence through an adversarial network, which allows for generalization without
necessarily requiring a growing number of training examples.

A common approach to account for systematic uncertainties in
a scientific context (e.g. in high energy physics)
is to take as fixed a classifier $f$ built from training data for a nominal
value $z_0$ of the nuisance parameter, and then propagate uncertainty by
estimating $p(f(x)|z)$ with a parametrized calibration procedure. Clearly, this
classifier is however not optimal for $z \neq z_0$. To overcome this issue, the
classifier $f$ is sometimes built instead on a mixture of training data
generated from several plausible values $z_0, z_1, \dots$ of the nuisance
parameter. While this certainly improves classification performance with respect
to the marginal model $p(X,Y)$, there is no reason to expect the resulting
classifier to be pivotal, as shown previously in Sec.~\ref{sec:toy}. As an
alternative, parametrized
classifiers~\citep{cranmer2015approximating,Baldi:2016fzo} directly take
(nuisance) parameters as additional input variables, hence ultimately providing
the most statistically powerful approach for incorporating the effect of
systematics on the underlying classification task.  As argued in
\citep{Neal:2007zz}, it is not obvious how such classifiers can be used on real
data since the correct value $z$ of the nuisance often remains unknown. This is
not an issue in the context of parameter
inference~\citep{cranmer2015approximating}, where nuisance parameters are
eliminated via optimization or marginalization. In practice, parametrized
classifiers  are also computationally expensive to build and evaluate. In
particular, calibrating their decision function, i.e. approximating
$p(f(x,z)|y,z)$ as a continuous function of $z$, remains an open challenge. By
contrast, constraining $f$ to be pivotal yields a classifier which may not be
optimal with respect to $Y|X,Z$, as discussed in Sec.~\ref{sec:theory}, but that
can otherwise be directly used in a wider range of applications, since the
dependence on the nuisance parameter $Z$ has already been eliminated. Moreover,
if $f$ is sufficiently close to being pivotal, then calibration only needs to be
carried out once.

@@

In terms of applications, the proposed solution can be used in any situation
where the training data may not be representative of the real data the
predictive model will be applied to in practice. In the scientific context, the
presence of systematic uncertainty can be incorporated by considering a family
of data generation processes, and it would be worth revisiting those scientific problems
that utilize machine learning in light of this technique. Moreover, the approach
also extends to cases where independence of the predictive model with respect to
observed random variables is desired, as in domain adaptation or fairness in
classification. Again, trying our algorithm on these use cases is certainly
worth investigating.

>>>>

\author{
  Gilles Louppe\\
  New York University\\
  \texttt{g.louppe@nyu.edu}\\
  \And
  Michael Kagan\\
  SLAC National Accelerator Laboratory\\
  \texttt{makagan@slac.stanford.edu}\\
  \And
  Kyle Cranmer\\
  New York University\\
  \texttt{kyle.cranmer@nyu.edu}
}

@@

\begin{abstract}
    % Many inference problems involve data generation processes that are not uniquely
    % specified or are uncertain in some way.
    In a scientific context, the presence of
    several plausible data generation processes is often associated to systematic uncertainties. Robust inference is possible if it is based on a
    pivot -- a quantity whose distribution does not depend on the unknown value of the
    (typically continuous) nuisance parameters that parametrize this family of
    generation processes. In this work,  we introduce a training procedure
    based on adversarial networks for enforcing the pivotal property on a predictive model.
    We derive theoretical results showing that the procedure tends
    towards a minimax solution corresponding to a predictive model that is both
    optimal and independent of the nuisance parameters (if that model exists) or
    for which one can tune a trade-off between accuracy and robustness. We
    demonstrate the effectiveness of this approach with a toy example and an example
    from particle physics.
\end{abstract}

@@

    \item $\smash{\hat\theta}_f$ maximizes the conditional entropy
    $H({Z|f(X;\theta_f)})$, since $H(Z|f(X;\theta)) \leq H(Z)$ from the properties of entropy. Note that this
    latter inequality holds for both the discrete and the differential definitions of entropy.
\end{itemize}
By assumption, the lower bound is active, thus we have $H(Z|f(X;\theta_f)) = H(Z)$
because of the second condition, which happens exactly when $Z$ and $f(X;\theta_f)$
are independent variables. In other words,  the
optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is also a pivotal
quantity.
\end{proof}

@@

\subsection{A toy example with a continous nuisance parameter}
\label{sec:toy}

@@

\begin{figure}
\centering
\begin{minipage}{.46\linewidth}
    \centering
        \includegraphics[width=\textwidth]{../figures/training.pdf}\vspace{-1em}
        \caption{Toy example. Training curves for ${\cal L}_f(\theta_f)$, ${\cal L}_r(\theta_f, \theta_r)$
                 and ${\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r)$.
                 Initialized with a pre-trained classifier $f$, adversarial training was performed for $200$ iterations, mini-batches of size $M=128$, $K=500$ and $\lambda=50$.}
        \label{fig:toy-training}
\end{minipage}
\hspace{.05\linewidth}
\begin{minipage}{.46\linewidth}
    \centering
        \includegraphics[width=\textwidth]{../figures/ams.pdf}\vspace{-1em}
        \caption{Physics example. Approximate median significance as a function of the decision threshold
                 on the output of $f$. At $\lambda=10$, trading
                 accuracy for independence to pileup
                 results in a net benefit in terms of statistical significance.}
        \label{fig:physics-ams}
\end{minipage}
\end{figure}

\subsection{High energy physics examples}
\label{sec:hep}

\paragraph{Binary Case} Experiments at high energy colliders like the LHC \citep{LHCMachine} are
searching for evidence of new particles beyond those described by the
Standard Model (SM) of particle physics. A wide array of
theories predict the existence of new massive particles that would decay to
known particles in the SM such as the $W$ boson. The $W$ boson is unstable and
can decay to two quarks, each of which produce collimated sprays of particles
known as jets.
%(collectively denoted $W\to jj$).
If the exotic particle
is heavy, then the $W$ boson will be moving very fast, and  relativistic effects
will cause the two jets from its decay to merge into a single `$W$-jet'. These
$W$-jets have a rich internal substructure (see e.g.
\citep{Altheimer:2012mn,Altheimer:2013yza}). However,
jets are also produced ubiquitously at high energy colliders through more
mundane processes in the SM, which leads to a challenging classification problem
that is beset with a number of sources of systematic uncertainty.
The classification challenge used here is common in jet substructure
studies (see e.g. \citep{Khachatryan:2014vla,ATL-PHYS-PUB-2015-033,wbosonATLAS}): we aim to distinguish normal jets produced copiously
at the LHC ($Y=0$) and from $W$-jets ($Y=1$) potentially coming from an exotic process.
We reuse the datasets used in
\citep{baldi2016jet}.
% For completeness, the jets are built with the
% anti-$k_T$ algorithm \citep{Cacciari:2008gp}
% with radius parameter $R=1.2$, and trimmed \citep{Krohn:2009th} with subjets built with
% the $k_T$ algorithm and parameter $f_{cut}=0.2$. The features used
% in the classifiers are: trimmed jet invariant mass, N-subjettiness
% $\tau_{21}^{\beta=1}$ \citep{nsub,Thaler2012}, and the energy correlation
% functions  $C_{2}^{\beta=1}$, $C_{2}^{\beta=2}$,
% $D_{2}^{\beta=1}$, $D_{2}^{\beta=2}$ \citep{Larkoski2013}.

@@

We do not expect that we will be able to find a function $f$ that simultaneously
minimizes the classification loss ${\cal L}_f$ and is pivotal. Thus, we need to optimize
the hyper-parameter $\lambda$ of Eqn.~\ref{eqn:vf-lambda} with respect to
a higher-level objective. In this case, the natural higher-level context is a
hypothesis test of a null hypothesis with no $Y=1$ events against an
alternate hypothesis that is a mixture of $Y=0$ and $Y=1$ events.
In the absence of systematic uncertainties, optimizing ${\cal L}_f$ simultaneously
optimizes the power of a classical hypothesis test in the Neyman-Pearson sense.
When we include systematic uncertainties we need to balance the
classification performance against the robustness to uncertainty in $Z$.
Since we are still performing a hypothesis test against the null, we only
wish to impose the pivotal property on $Y=0$ events. To this end,
we use as a higher level objective the Approximate Median Significance (AMS), which is a natural generalization
of the power of a hypothesis test when systematic uncertainties are taken into account
(see Eqn.~20 of \cite{adam2014higgs}).

@@

\paragraph{Continous Case} 
%Our method also applies to cases where one desires the predictive model to be independent of 
%some other observed random variables, as in domain adaptation or fairness in
%classification. Importantly, our approach is able to enforce fairness over
%continuous attributes.
Recently, an independent group has used our approach to learn jet classifiers 
that are independent of the jet mass, which is a continuous attribute~\citep{Shimmin:2017mfk}. 
The results of their studies show that the adversarial training strategy works very well for 
real-world problems with continuous attributes, thus enhancing the sensitivity of searches for 
new physics at the LHC.

@@

%While our contribution may first appear as a minor extension, most
Most related
work is based on the strong and limiting assumption that $Z$ is a binary
random variable (e.g., $Z=0$ for the source domain, and
$Z=1$ for the target domain). In particular, \citep{pan2011domain,gong2013connecting,baktashmotlagh2013unsupervised,zemel2013learning,ganin2014unsupervised,ajakan2014domain,EdwardsS15,louizos2015variational}
are all based on the minimization of some form of divergence between the two distributions of
$f(X)|Z=0$ and $f(X)|Z=1$. For this reason, 
%and in contrast with our procedure,
these works cannot directly be generalized to non-binary or
continuous nuisance parameters, both from a practical and theoretical point of view.
Notably, \cite{kamishima2012fairness} enforces fairness through
a prejudice regularization term based on empirical estimates of $p(f(X)|Z)$.
While this approach is in principle sufficient for handling non-binary
nuisance parameters $Z$, it requires accurate
empirical estimates of $p(f(X)|Z=z)$ for all values $z$, which quickly becomes
impractical as the cardinality of $Z$ increases. By contrast, our approach models the conditional
dependence through an adversarial network, which allows for generalization without
necessarily requiring a growing number of training examples.

A common approach to account for systematic uncertainties in
a scientific context (e.g. in high energy physics)
is to take as fixed a classifier $f$ built from training data for a nominal
value $z_0$ of the nuisance parameter, and then propagate uncertainty by
estimating $p(f(x)|z)$ with a parametrized calibration procedure. Clearly, this
classifier is however not optimal for $z \neq z_0$. To overcome this issue, the
classifier $f$ is sometimes built instead on a mixture of training data
generated from several plausible values $z_0, z_1, \dots$ of the nuisance
parameter. While this certainly improves classification performance with respect
to the marginal model $p(X,Y)$, there is no reason to expect the resulting
classifier to be pivotal, as shown previously in Sec.~\ref{sec:toy}. As an
alternative, parametrized
classifiers~\citep{cranmer2015approximating,Baldi:2016fzo} directly take
(nuisance) parameters as additional input variables, hence ultimately providing
the most statistically powerful approach for incorporating the effect of
systematics on the underlying classification task.  
%As argued in
%\citep{Neal:2007zz}, it is not obvious how such classifiers can be used on real
%data since the correct value $z$ of the nuisance often remains unknown. This is
%not an issue in the context of parameter
%inference~\citep{cranmer2015approximating}, where nuisance parameters are
%eliminated via optimization or marginalization. 
In practice, parametrized
classifiers  are also computationally expensive to build and evaluate. In
particular, calibrating their decision function, i.e. approximating
$p(f(x,z)|y,z)$ as a continuous function of $z$, remains an open challenge. By
contrast, constraining $f$ to be pivotal yields a classifier 
%which may not be
%optimal with respect to $Y|X,Z$, as discussed in Sec.~\ref{sec:theory}, but 
that can 
%otherwise 
be directly used in a wider range of applications, since the
dependence on the nuisance parameter $Z$ has already been eliminated. 
%Moreover, if $f$ is sufficiently close to being pivotal, then calibration only needs to be
%carried out once.

@@

In terms of applications, the proposed solution can be used in any situation
where the training data may not be representative of the real data the
predictive model will be applied to in practice. In the scientific context, the
presence of systematic uncertainty can be incorporated by considering a family
of data generation processes, and it would be worth revisiting those scientific problems
that utilize machine learning in light of this technique. Moreover, the approach
also extends to cases where independence of the predictive model with respect to
observed random variables is desired, as in domain adaptation or fairness in
classification (e.g. \citep{Shimmin:2017mfk}).
% not sure this is a great sentence to end with, sounds like a sales pitch
% Again, trying our algorithm on these use cases is certainly
%worth investigating.
